\mbox{Specifying Imperative Higher-Order Programs}

### 2009-10-05 papers/esop2010/main.tex
advancing on the paper

\newcommand{\wpre}{{\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{\langle #1 \rangle}
\newcommand{\lift}[1]{\langle #1 \rangle}
\newcommand{\alist}[1]{\overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\ocaml}{Ocaml}
\newcommand{\correct}[1]{c(#1)}

@@

\author{Johannes Kanig \and Jean-Christophe Fill\^atre}
\institute{
Laboratoire de Recherche en Informatique, Univ Paris-Sud, CNRS, Orsay F-91405\\
INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
}
%TODO affiliation
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} gives an introduction to the described language as
well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
details about the programming language and the weakest precondition calculus.
In section~\ref{sec:implem} we speak about the implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

\section{The general picture and a motivating example}
\label{sec:genpic}
%TODO Hoare logic = sharing between logic and prog language + alias free
also one or two examples in the described language

characteristics : 
\begin{itemize}
  \item Hoare logic
  \item logic and program share values  
  \item Type and effect system
  \item Region typing to track references
  \item the wp needs absence of aliasing to produce good POs
\end{itemize}

@@

technical presentation:
\begin{itemize}
  \item  notion of substitution - correspondence χ κ
  \item semantics is standard (goes to appendix )
\end{itemize}

\subsection{Syntax}
\begin{figure}[htpb]
\begin{eqnarray*}
  x & & \text{Program Variables}\\
  α & & \text{Type Variables}\\
  e & & \text{Effect Variables}\\
  ρ & & \text{Region Variables}\\
  χ &::=& α \sep e \sep ρ \\
  t,p,q &::=& x~[\alist{κ}] \sep λx:τ.t \sep \recml~f~(x:τ).\{p\}t\{q\} \sep t~t\sep \\ 
  & & \letml~ x~[\alist{χ}] = t~\inml~t \sep \ifml~t~\thenml~t~\elseml~t\\
  τ &::=& α \sep τ -> τ \sep τ ->^ε τ \sep τ \times τ \sep τ~\refml_ρ \sep
  \efft{ε} \\
  ε &::=& (ρ \sep e)*\\
  κ &::=& τ \sep ε \sep ρ
\end{eqnarray*}
  \caption{Syntax}
  \label{fig:syntax}
\end{figure}

Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
with \letml-bindings, where there are two different λ-bindings. The first one
($λx:τ.t$) describes a pure, non-recursive binding, and the second one
($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
functions; $p$ and $q$ in this term are {\em specifications}, which are
of the same syntax as programs, but are differently typed (typing will be
discussed in the next section). The term $p$ corresponds to the {\em
precondition}, $q$ to the {\em postcondition} of the function. The analogy to
Hoare triples is of course obvious. As usual, there are also variables,
\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
level can be generalized at \letml-boundaries and have to be instantiated when
the corresponding program variable is used.

@@

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
  \efft{ε} -> τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

\paragraph{Value typing.}
Let us first look at the common part concerning values. We define a relation
$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
is of type $τ$''. Values cannot have any effect.  Variables and pure
abstractions are typed in the usual way; type instantiations at variable
occurences are substituted in the type. The rule {\sc REC} simply states that
the body of the function can use the argument $x$ as well as the function $f$.
The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
wrt. the typing relation $|-_l$ described later on. Finally, pure applications
between values are considered as values as well, as they don't have any
effect.

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
  } \and
  { \inferrule*[Left=Sub]
    { Γ |- t : τ', ε \\ ε \subseteq ε' }
    { Γ |- t : τ, ε' }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

@@

All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produe
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

The typing of annotations is basically an extension of the typing for values;
logic terms never have any effect. To ensure this, we need to forbid calling
effectful functions in the logic. One way is to simply omit the typing rule
for effectful function calls. There is, however, another way which also
enables us to inspect the pre/ and postcondition of a function. To this end,
let us introduce the operation of {\em lifting} a type to the logic level (see
Fig.~\ref{fig:typelift}).
\begin{figure}[htpb]
  \begin{eqnarray*}
    \ceil{α} &=& α\\
    \ceil{τ~\refml_ρ} &=& \ceil{τ}~\refml_ρ\\
    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
    \ceil{τ \times τ'} &=& \ceil{τ} \times \ceil{τ'}\\
    \ceil{\efft{ε}} &=& \efft{ε}\\
    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)
  \end{eqnarray*}
  \caption{Lifting types}
  \label{fig:typelift}
\end{figure}
This translation traverses the type and replaces all effectful types by a
tuple type which describes a tuple of a pre- and a postcondition. The
precondition takes as an argument the argument of the function as well as the
part of the store, at the moment of the function call, which corresponds to
the effect of the function. The postcondition's two first arguments are the
same, but in addition it expects the store {\em at the exit of the function},
as well as the function's return value.

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
\propml -> \propml \end{equation*} Actually, when equality is given, all
logical operations (even quantification) can be defined in the
logic~\cite{andrews86}.  In the remainder of the paper, we suppose
predefined the usual logical connectives conjunction, disjunction,
implication and equivalence, as well as universal and existential
quantification. We also assume all the arithmetic operations as well as the
constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs.

%TODO pre and post
% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

\begin{figure}[htpb]
  \begin{mathpar}
    {\inferrule*[Left=Var]
      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
      {Γ |- x : τσ }
    } \\
    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
    }
  \end{mathpar}
  \caption{A modified and a new rule for aliasing}
  \label{fig:aliasing}
\end{figure}

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
%TODO
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}

which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a composite substitution is compatible
%TODO atomic and composite
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.

>>>>

\newcommand{\wpre}{{\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\preml}{{\bf pre}}
\newcommand{\postml}{{\bf post}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{\langle #1 \rangle}
\newcommand{\lift}[1]{\langle #1 \rangle}
\newcommand{\alist}[1]{\overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\unitml}{{\bf unit}}
\newcommand{\intml}{{\bf int}}
\newcommand{\coq}{Coq}
\newcommand{\who}{Who}
\newcommand{\ocaml}{Ocaml}
\newcommand{\correct}[1]{c(#1)}

@@

\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
Fill\^atre\inst{2,}\inst{1}}
\institute{
INRIA Saclay - Île-de-France, ProVal, Orsay, F-91894 \and
 LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
}
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

%TODO clearly state possibilities/limitations of the system
% * one can store everything in refs
% * but aliasing restriction (refs in refs, lists of refs etc)

@@

\paragraph{Outline.}
Section~\ref{sec:genpic} gives an introduction to the described language as
well as an example. Sections~\ref{sec:syntax} and \ref{sec:wp} give the
details about the programming language and the weakest precondition calculus.
In section~\ref{sec:examples} we speak about the implementation and show some
examples. Finally, section~\ref{sec:related} discusses related work.

\section{The general picture and a motivating example}
\label{sec:genpic}

Consider the following first-order program\footnote{Throughout the document,
we use \ocaml~Syntax for programs.} that accumulates the sum of the contents
of some array $a$ in some global reference $c$, using an integer $i$ as
a counter:
\begin{ocaml}
  let sum () = 
    for i = 0 to Array.length a - 1 do
      c := !c + a.(i)
\end{ocaml}
This simple  can be annotated in a standard Hoare logic (for
example the system of \cite{Filliatre00a}) as follows:
\begin{who}
  let sum () =
  {!!c = 0}
    for i = 0 to length !a - 1 do
    { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} }

      c := !c + get !a i

      {!!c = {{{ $\sum_0^{ {\tt length~!a} - 1}$ get !!a i }}} }
\end{who}

Here, assuming we model the array in the programming language by a reference
to a functional array, with functions {\tt length} and {\tt get} with the
obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
instead of the usual {\tt !}. In this case, our program is a {\tt for} loop,
so it needs an {\em invariant annotation}, indicating which relation is
preserved by the loop body. The equalities that need to be proved are very
simple. Also, it is obvious that the program modifies exactly the references
$r$ and $c$, and this fact can be formally confirmed by an effect analysis.

A common reaction of a functional programmer to a program like
$\mathit{sum}$ is to factor out the structure of the recursion, which
may be useful in other situations, in a higher-order function. The
more specific program is then recovered by function application. In this case,
the more general function is of course simply {\tt Array.iter}:
\begin{ocaml}
  let iter f a = 
    for i = 0 to Array.length a - 1 do
      f a.(i)

  let sum () =
    iter (fun n -> c := !c + n) a
\end{ocaml}
It is obvious that the two programs do exactly the same
thing. However, how do we prove the correctness of the second one? One
possibility is to fix the future usage of ${\tt iter}$ by
specifying what exactly is the specification of its argument {\tt f}:
\begin{who}
  let iter 
    (f : (x : int) ->{c} 
        { !!c = {{{ $\sum_0^{i-1}$ get !!a i }}} } 
        unit 
        { !!c = {{{ $\sum_0^{length !a - 1}$ get !!a i }}} } ) 
    (a : int array) =
    ...
\end{who}
Here we impose precisely the specification of {\tt f}, and thus we can
derive that {\tt sum} computes the right value,
using the same method as for the first-order case before. Note
that, to be able to apply an effect analysis, we have to specify that
$f$ modifies $c$, by giving it the type $\intml->^c\unitml$.

Of course, this approach is not very satisfactory. While the program achieves
the separation of the iterating function ({\tt iter}) and the combining
function (the anonymous function), the specification does not do this at all;
as specified,  {\tt iter} can basically only be applied to the same function,
and can only compute a certain sum. If one wants to use {\tt iter} in another
context, one has to copy the code, give a new specification adapted to the
application and prove this new specification again. In addition, in order to
be able to write the specification at all, we had to specialize the type of
{\tt iter} to integer arrays, while its \ocaml\ counterpart is of course
polymorphic. Said otherwise, this approach is not modular at all.

There are two obstacles related to higher-order functions to be
overcome in order to achieve modularity on the level of specification
in the presence of side effects. We must be able to refer to the
effect of a function parameter, and we must be able to refer to the
specification (pre- and postconditions) of a function parameter, in a
generic way.

The first obstacle has a well known solution using a type and effect
system~\cite{talpin94taed} with effect variables: effect polymorphism
does exactly what we want here. Using effect polymorphism, we can say
that {\tt f} (the argument of {\tt iter}) has type
$\intml->^e\unitml$, where $e$ is an {\em effect variable}, which
amounts to say that {\tt f} has a {\em certain} effect $e$. The total
effect of {\tt iter} is then $e$, meaning {\tt iter}
modifies whatever {\tt f} modifies.

A solution to the second problem concerning specifications consists in
representing a program function in the logic as the pair of its pre- and
postcondition. This is an old idea, and it has recently been applied to purely
functional programs by Régis-Gianas and
Pottier~\cite{regis-gianas-pottier-08}.

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Th so-called
two-state postconditions are a way to get rid of most uses of auxiliary
variables. However, the pre- and postcondition of {\tt f} may not refer to
arbitrary parts of the state, but only the part {\tt f} reads or modifies.
With these conventions, one can now write a generic specification for the {\tt
iter} function, see ~Fig.~\ref{fig:iter}.

\begin{figure}[tbp]
\begin{who}[lines]
let iter [e] (a : 'a farray ref) (inv : <e> -> int ->  prop)
  (f : 'a ->{e} unit) =
  { inv cur 0 /\ forall (i:int). 0 <= i /\ i < length !!a ->
    forall (m : <e>).  inv m i -> pre f (get i !!a) m /\
    forall (n : <e>).  post f (get i !!a) m n () -> inv n (i+1)
  }
  for index = 0 to length !a - 1 do
    { inv cur index }
    f (get index !a)
  done
  {inv cur (length !!a) }
\end{who}
\caption{The specification of the {\tt iter} function}
  \label{fig:iter}
\end{figure}

There are several new elements here. First of all, we use effect polymorphism
to introduce an effect variable {\tt e} standing for the effect of {\tt f}.
Also, {\tt iter} has an additional argument {\tt inv}, the {\em invariant},
taking as argument the portion of the state modified by f and the integer
indicating the current iteration count. This is a logical argument, which
means that it cannot be used in the program. Also, {\tt inv} is a {\em pure}
function, which is indicated by the arrows without effect annotation. The
postcondition is now quite clear: we simply want to establish the invariant
for the final state (the keyword {\tt cur} for the current state of the
postcondition) and the length of the array, indicating that we have iterated
over the entire array. It is also easy to see that the invariant {\tt inv} is
precisely the invariant of the {\tt for} loop. However, the precondition is a
bit complicated: it states that we need the invariant to hold initially (for
the current state of the precondition and index $0$) and we need the invariant
to be preserved by the function call. This is expressed by quantifying over
two states, saying that the invariant has to imply the precondition of {\tt f}
and the postcondition has to imply the invariant for the next array index.
With a bit of syntactic sugar, this could be expressed much more concisely
using a Hoare triple:
\begin{who}
  { inv cur i} f (get i !!a) {inv cur (i+1)}
\end{who}

The {\tt iter} function and its specification raise a question: is the
specification correct if {\tt f} modifies the array {\tt a}, which is required
for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
invariant.  This is an instance of the {\em aliasing problem} (see, for
example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
does not support aliasing. In this paper we make the same choice and disallow
aliasing in the programming language we introduce. Under this assumption, the
specification we gave is correct.

Entirely excluding aliasing renders our system inappropriate for
reasoning about shared mutable structures. However, it is interesting
to see that our alias exclusion restriction, in many cases,
corresponds to a natural condition imposed on higher-order functions
that operate on some data structure. It is a common
requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
functions, for example, that the functional argument must not modify
the structure to be iterated on. In our system, this fact follows from
the restriction on effect instantiations, as will be seen in
section~\ref{sec:syntax}. In the example, the array {\tt a}
of the iterating function {\tt iter} must not be modified by its
argument.

@@

%TODO  notion of substitution - correspondence χ κ
\subsection{Syntax}
\begin{figure}[tpb]
\begin{eqnarray*}
  x & & \text{Program Variables}\\
  α & & \text{Type Variables}\\
  e & & \text{Effect Variables}\\
  ρ & & \text{Region Variables}\\
  χ &::=& α \sep e \sep ρ \\
  t,p,q &::=& x~[\alist{κ}] \sep λx:τ.t \sep \recml~f~(x:τ).\{p\}t\{q\} \sep t~t\sep \\ 
  & & \letml~ x~[\alist{χ}] = t~\inml~t \sep \ifml~t~\thenml~t~\elseml~t\\
  τ &::=& α \sep τ -> τ \sep τ ->^ε τ \sep τ \times τ \sep τ~\refml_ρ \sep
  \efft{ε} \\
  ε &::=& (ρ \sep e)*\\
  κ &::=& τ \sep ε \sep ρ
\end{eqnarray*}
  \caption{Syntax}
  \label{fig:syntax}
\end{figure}

Our programming language (Fig.~\ref{fig:syntax}) is basically the λ-calculus
with \letml-bindings, where there are two different λ-bindings. The first one
($λx:τ.t$) describes a pure, non-recursive binding, and the second one
($\recml~f~(x:τ).\{p\}t\{q\}$) describes recursive, potentially impure
functions; $p$ and $q$ in this term are {\em specifications}, which are
of the same syntax as programs, but are differently typed (typing will be
discussed in the next subsection). The term $p$ corresponds to the {\em
precondition}, $q$ to the {\em postcondition} of the function. The analogy to
Hoare triples is of course obvious. As usual, there are also variables,
\letml-bindings and an \ifml-\thenml-\elseml-construct. Variables on the type
level can be generalized at \letml-boundaries and have to be instantiated when
the corresponding program variable is used.

@@

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
  \efft{ε} -> τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

\paragraph{Value typing.}
Let us first look at the common part concerning values. We define a relation
$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
is of type $τ$''. Values cannot have any effect.  Variables and pure
abstractions are typed in the usual way; type instantiations at variable
occurrences are substituted in the type. The rule {\sc REC} simply states that
the body of the function can use the argument $x$ as well as the function $f$.
The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
wrt. the typing relation $|-_l$ described later on. Finally, pure applications
between values are considered as values as well, as they don't have any
effect.

\begin{figure}[tpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
  } \and
  { \inferrule*[Left=Sub]
    { Γ |- t : τ', ε \\ ε \subseteq ε' }
    { Γ |- t : τ, ε' }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

@@

All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produce
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

@@

The typing of annotations is basically an extension of the typing for values;
logic terms never have any effect. To ensure this, we need to forbid calling
effectful functions in the logic. One way is to simply omit the typing rule
for effectful function calls. There is, however, another way which also
enables us to inspect the pre- and postcondition of a function. To this end,
let us introduce the operation of {\em lifting} a type to the logic level (see
Fig.~\ref{fig:typelift}).
\begin{figure}[tpb]
  \begin{eqnarray*}
    \ceil{α} &=& α\\
    \ceil{τ~\refml_ρ} &=& \ceil{τ}~\refml_ρ\\
    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
    \ceil{τ \times τ'} &=& \ceil{τ} \times \ceil{τ'}\\
    \ceil{\efft{ε}} &=& \efft{ε}\\
    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)
  \end{eqnarray*}
  \caption{Lifting types}
  \label{fig:typelift}
\end{figure}
This translation traverses the type and replaces all effectful types by a
tuple type which describes a tuple of a pre- and a postcondition. The
precondition takes as an argument the argument of the function as well as the
part of the store, at the moment of the function call, which corresponds to
the effect of the function. The postcondition's two first arguments are the
same, but in addition it expects the store {\em at the exit of the function},
as well as the function's return value.

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjunction is just a function: \begin{equation*} /\ :  \propml ->
\propml -> \propml \end{equation*} Actually, when equality is given, all
logical operations (even quantification) can be defined in the
logic~\cite{andrews86}.  In the remainder of the paper, we suppose
predefined the usual logical connectives conjunction, disjunction,
implication and equivalence, as well as universal and existential
quantification. We also assume all the arithmetic operations as well as the
constructor $(t,t)$  and the accessors $fst$ and $snd$ for pairs. We will also
use the names \preml\ and \postml\ for $fst$ and $snd$, respectively.

% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

\begin{figure}[tpb]
  \begin{mathpar}
    {\inferrule*[Left=Var]
      {Γ(x) = ∀\alist{χ}.τ \\ σ = [\alist{χ}|->\alist{κ}] \\ σ \sim τ }
      {Γ |- x : τσ }
    } \\
    { \inferrule*[Left=LetRef, Right={ρ \textup{fresh}, $ρ\notin τ$}]
      {Γ |-_v t1 : τ' \\ Γ, x : τ'~\refml_ρ |- t2 : τ, ε }
      {Γ |- \letml~x = ref~ t1 ~\inml~ t2 : τ, ε \setminus ρ}
    }
  \end{mathpar}
  \caption{A modified and a new rule for aliasing}
  \label{fig:aliasing}
\end{figure}

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}
which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} region substitution (whose domain contains only one
variable) $[ρ|->ρ']$. It is compatible with any type which does not contain
$ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2 e'$] is compatible with
any type which does not contain $ ρ1 , ρ2 $ nor $e'$.  Expressed more
generally, an atomic substitution is compatible with a type if the type does
not contain any region or effect variables which are in the range of the
substitution. Finally, a {\em composite} substitution (whose domain contains more
than one variable) is compatible
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.
